---
title: "Interpreting Coefficients"
author: "Nicholas Reich, adapted from Project MOSAIC"
date: "February 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

We will load the data set JAMA-HR.csv which contains data from a report in the Journal of the American Medical Association.

Each case is an individual person. The variables are 

 - Temp: body temperature in degrees F 
 - HR: heart rate in beats per minute
 - Sex: coded as F and M

Here are some initial commands to load and look at the dataset: 
```{r, message=FALSE}
jama = fetch::fetchData("JAMA-HR.csv")
nrow(jama)
head(jama)
```

## Models
Here are several different model outputs for you to consider. Recall that the `coef()` function returns a named vector with the coefficients from the fitted model.

```{r}
mod1 = lm(HR ~ 1, data=jama) 
coef(mod1)
mod2 = lm(HR ~ Temp, data=jama) 
coef(mod2)
mod3 = lm(HR ~ Sex, data=jama) 
coef(mod3)
mod4 = lm(HR ~ Temp + Sex, data=jama) 
coef(mod4)
mod5 = lm(HR ~ Sex * Temp, data=jama) 
coef(mod5)
```

(b) What is the mean heart rate just for the men?
(c) What are the units of the coefficient -166.285 in Model 2?
(d) What are the units of the coefficient 2.443 in Model 2?
(e) What is the meaning for heart rate of the coefficient 2.443 in Model 2?
(f) Using Model 4, what is the model value for the heart rate of a woman with temperature 100deg?
(g) The coefficient SexM in Model 4 is negative. Describe what this indicates in terms of everyday language.
(h) Using Model 5, what is the model value of the heart rate for a man with temperature 100deg?
(i) What is the slope of a graph (HR vs Temp) of the model function for women in Model 5? Include units.
(j) What is the slope of a graph (HR vs Temp) of the model function for men in Model 5? Include units.
